Staged Abstract Interpreters

### 2019-04-06 paper/20_prelim.tex
Grammar Prelim

We consider a call-by-value $\lambda$-calculus in direct-style, extended
with numbers, arithmetic, recursions and conditionals. Other effectful features
such as assignments can also be supported readily.
%In Section~\ref{cases_imp}, we will add more imperative features to the
language. Since we are mostly
interested in analyzing the dynamic behaviors of the program, we disguise any
static semantics and type system. We also assume that input programs are
well-typed and all variables are distinct. The abstract syntax is shown as
follows:

@@

\iffalse
We will give the concrete semantics using a big-step definitional
interpreter. The interpreter is a recursive function that takes the program AST,
environment and store, and returns the evaluated value and the accompanied
store. The environment is a mapping from identifiers to addresses, and the store
is a mapping from addresses to values. We use the store to model recursion and
mutation in concrete semantics; it is also useful for polyvariant analysis. This
environment-and-store-passing style big-step interpreter is standard and can
also be obtained by refunctionalizing \cite{DBLP:conf/ppdp/AgerBDM03,
Wei:2018:RAA:3243631.3236800} a small-step CESK machine
\cite{DBLP:conf/popl/FelleisenF87}.
\fi

@@

A monad is a type constructor @M[_]: * -> *@ with two operations, often called
@return@ and @bind@. Informally, @return@ wraps a value into the monad @M@, and
@bind@ unwraps the monadic value and transforms it to a new monadic value.
Pragmatically in Scala, we define a monad type class using trait @Monad@ (Figure
\ref{fig:monad}), where it declares the @pure@ \footnote{We elect to use
\texttt{pure} as the name, since \texttt{return} is a keyword in Scala and
\texttt{unit} is a built-in function in LMS.} and @flatMap@ operation. The trait
itself takes the monad type @M[_]@ as argument, which is a higher-kinded type
that takes a type and returns a type. The method @pure@ promotes values of type
@A@ to values of type @M[A]@. The monadic @bind@ operation is usually called
@flatMap@ in Scala, which takes a monad-encapsulated value of type @M[A]@, a
function of @A => M[B]@ and returns values of type @M[B]@.

@@

Similar to Haskell's @do@-notation, Scala provides special syntactic support for
monadic operations through @for@-comprehension.
For example, an object of @List[A]@ is an instance of @List@ monad, where @A@ is the element type. 
Then to compute the the Cartesian product of two lists of numbers, we can use Scala's
@for@-comprehension syntax.

@@

Note that here the monadic object @List[_]@ encapsulates the data internally.
Therefore it only exposes the simplified version of @flatMap@, where the
monad @M[A]@ is not introduced as a function argument. The trait @MonadOps@
(Figure \ref{fig:monadops}) defines the simplified version of monadic
operations that are necessary for @for@-comprehension. 
The conversion between @Monad@ and @MonadOps@ can be done by using implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax and monad
transformers such as @ReaderT@, @StateT@, and @ListT@ to write our interpreters.
The implementation of monads and monad transformers essentially borrows the
ground-truth from Haskell.

@@

\paragraph{Monadic Operations} We define the return type of the interpreter as
@Ans@, which is an arbitrary monad type @AnsM[_]@ wrapping the type @Value@. 
As mentioned in Section ~\ref{monadscala}, to use the @for@-comprehension
syntax, certain constraints have to be posed onto the type @AnsM@. Here, we use a
structural type @MonadOps@ to require @AnsM@ to at least implement @map@ and
@flatMap@. It is worth noting that @MonadOps@ takes another type parameter
@R[_]@ as binding-time; accordingly the outer @R@ defined in the trait is passed
to @MonadOps@. Inside of @MonadOps@, @R[_]@ wraps the data types @A@ and @B@ that are
encapsulated by the monad, but not the monad type @M@ itself. When acting as
compilers, we will also replace the monads to the ones that work on staged
values.

We also defines several methods to obtain and modify the environment and store.
These methods return monadic values of type @AnsM[_]@, which encapsulate the 
environment or store, or simply a @Unit@ value for effects.
For example, @local_env@ takes a value @ans@ of type @Ans@ and an environment 
which will be installed when evaluating @ans@.

@@

\paragraph{Primitive Operations} Next we define several primitive operations.
Two versions of @alloc@ are declared. The first one takes a store and an
identifier, and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is also wrapped
by @R[_]@. The other one would simply wrap the address with our monadic type
@AnsM[_]@.
\begin{lstlisting}
  def alloc(σ: R[Store], x: Ident): R[Addr];  def alloc(x: Ident): AnsM[Addr]
\end{lstlisting}

@@

Since the interpreter is written in open-recursive style, we declare an abstract
combinator @fix@ used to close the recursion. For concrete-interpretation instantiation, it
works like the Y combinator; for abstract-interpretation instantiation, we will
instrument the interpreter by defining a memorized version of @fix@ to ensure
termination. Finally, a top-level wrapper @run@ is declared; the return type
@Result@ depends on what kind of monad we will be using, thus is also an abstract type.

@@

\paragraph{Unstaged Monads}
The environment and store can be modeled by reader effect and state effect,
i.e., a Reader monad and a State monad. We combine them using monad transformers
@ReaderT@ and @StateT@.
In other word, we instantiate the @AnsM@ monad as a stack of @ReaderT@ and @StateT@
transformers\footnote{The question mark syntax is a kind projector
  \cite{kindprojector}, thus \texttt{StateT[IdM,Store,?]} is equivalent to \newline
  \texttt{(\{type M[T]=StateT[IdM,Store,T]\})\#M}}, where the @ReaderT@ is
parameterized by type @Env@, @StateT@ is parameterized by type @Store@, and the
inner-most monad @IdM@ is just an identity monad.

@@

Then, the monadic operations that manipulate the environment and store are
simply constructing the proper monads and lifting it to the top-level of our
moand stack. To modify the store, for example, we may have a @StateT@ monad that
transforms the current store $\sigma$ to @σ + αv@, which results in
@StateT[IdM, Store, Unit]@, and then lift this @StateT@ value to @ReaderT@,
i.e., @AnsM[Unit]@.
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ + αv))
\end{lstlisting}

\paragraph{Primitive Operations}
Other primitive operations can be implemented straightforwardly, for example,
@alloc@ simply takes the size of current store and produce the successor number.
We elide most of them, but describe a little bit on how do we handle functions
and applications as they will be very different when staging and abstract
interpretation is involved. The method @close@ takes a lambda term and an
environment, then create a closure value @CloV@.

@@

To apply a function, the method @ap_clo@ takes a function value and an arugment
value, extracts the lambda term and environment enclosed in the function value,
and evaluates the body expression of that lambda term under the new environment and store.

@@

With other primitive operations implemented, we can implement the top-level
@run@ method, where $\rho_0$ and $\sigma_0$ are initial empty environment and
store.

>>>>

We consider a call-by-value $\lambda$-calculus in direct-style, extended
with numbers, arithmetic, recursions, and conditionals. Other effectful features
such as assignments can also be supported readily.
%In Section~\ref{cases_imp}, we will add more imperative features to the
language. Since we are mostly
interested in analyzing the dynamic behaviors of the program, we disguise any
static semantics and type system. We also assume that input programs are
well-typed and all variables are distinct. The abstract syntax is shown as
follows:

@@

\iffalse
We will give the concrete semantics using a big-step definitional
interpreter. The interpreter is a recursive function that takes the program AST,
environment, and store, and returns the evaluated value and the accompanying
store. The environment is a mapping from identifiers to addresses, and the store
is a mapping from addresses to values. We use the store to model recursion and
mutation in concrete semantics; it is also useful for polyvariant analysis. This
environment-and-store-passing style big-step interpreter is standard and can
also be obtained by refunctionalizing \cite{DBLP:conf/ppdp/AgerBDM03,
Wei:2018:RAA:3243631.3236800} a small-step CESK machine
\cite{DBLP:conf/popl/FelleisenF87}.
\fi

@@

A monad is a type constructor @M[_]: * -> *@ with two operations, often called
@return@ and @bind@. Informally, @return@ wraps a value into the monad @M@, and
@bind@ unwraps the monadic value and transforms it into a new monadic value.
Pragmatically in Scala, we define a monad type class using trait @Monad@ (Figure
\ref{fig:monad}), where it declares the @pure@ \footnote{We elect to use
\texttt{pure} as the name, since \texttt{return} is a keyword in Scala and
\texttt{unit} is a built-in function in LMS.} and @flatMap@ operation. The trait
itself takes the monad type @M[_]@ as argument, which is a higher-kinded type
that takes a type and returns a type. The method @pure@ promotes values of type
@A@ to values of type @M[A]@. The monadic @bind@ operation is usually called
@flatMap@ in Scala, which takes a monad-encapsulated value of type @M[A]@, a
function of @A => M[B]@ and returns values of type @M[B]@.

@@

Similar to Haskell's @do@-notation, Scala provides special syntactic support for
monadic operations through @for@-comprehension.
For example, an object of @List[A]@ is an instance of @List@ monad, where @A@ is the element type. 
Then to compute the Cartesian product of two lists of numbers, we can use Scala's
@for@-comprehension syntax.

@@

Note that here the monadic object @List[_]@ encapsulates the data internally.
Therefore it only exposes the simplified version of @flatMap@, where the
monad @M[A]@ is not introduced as a function argument. The trait @MonadOps@
(Figure \ref{fig:monadops}) defines the simplified version of monadic
operations that are necessary for @for@-comprehension. 
The conversion between @Monad@ and @MonadOps@ can be done by using the implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax and monad
transformers such as @ReaderT@, @StateT@, and @ListT@ to write our interpreters.
The implementation of monads and monad transformers essentially borrows the
ground-truth from Haskell.

@@

\paragraph{Monadic Operations} We define the return type of the interpreter as
@Ans@, which is an arbitrary monad type @AnsM[_]@ wrapping the type @Value@. 
As mentioned in Section ~\ref{monadscala}, to use the @for@-comprehension
syntax, certain constraints have to be added on the type @AnsM@. Here, we use a
structural type @MonadOps@ to require @AnsM@ to at least implement @map@ and
@flatMap@. It is worth noting that @MonadOps@ takes another type parameter
@R[_]@ as binding-time; accordingly the outer @R@ defined in the trait is passed
to @MonadOps@. Inside of @MonadOps@, @R[_]@ wraps the data types @A@ and @B@ that are
encapsulated by the monad, but not the monad type @M@ itself. When acting as
compilers, we will also replace the monads to the ones that work on staged
values.

We also define several methods to obtain and modify the environment and store.
These methods return monadic values of type @AnsM[_]@, which encapsulate the 
environment or store, or simply a @Unit@ value for effects.
For example, @local_env@ takes a value @ans@ of type @Ans@ and an environment 
which will be installed when evaluating @ans@.

@@

\paragraph{Primitive Operations} Next we define several primitive operations.
Two versions of @alloc@ are declared. The first one takes a store and an
identifier and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is also wrapped
by @R[_]@. The other one would simply wrap the address with our monadic type
@AnsM[_]@.
\begin{lstlisting}
  def alloc(σ: R[Store], x: Ident): R[Addr];  def alloc(x: Ident): AnsM[Addr]
\end{lstlisting}

@@

Since the interpreter is written in an open-recursive style, we declare an abstract
combinator @fix@ used to close the recursion. For concrete-interpretation instantiation, it
works like the Y combinator; for abstract-interpretation instantiation, we will
instrument the interpreter by defining a memorized version of @fix@ to ensure
termination. Finally, a top-level wrapper @run@ is declared; the return type
@Result@ depends on what kind of monad we will be using, this is also an abstract type.

@@

\paragraph{Unstaged Monads}
The environment and store can be modeled by reader effect and state effect,
i.e., a Reader monad and a State monad. We combine them using monad transformers
@ReaderT@ and @StateT@.
In other words, we instantiate the @AnsM@ monad as a stack of @ReaderT@ and @StateT@
transformers\footnote{The question mark syntax is a kind projector
  \cite{kindprojector}, thus \texttt{StateT[IdM,Store,?]} is equivalent to \newline
  \texttt{(\{type M[T]=StateT[IdM,Store,T]\})\#M}}, where the @ReaderT@ is
parameterized by type @Env@, @StateT@ is parameterized by type @Store@, and the
inner-most monad @IdM@ is just an identity monad.

@@

Then, the monadic operations that manipulate the environment and store are
simply constructing the proper monads and lifting it to the top-level of our
monad stack. To modify the store, for example, we may have a @StateT@ monad that
transforms the current store $\sigma$ to @σ + αv@, which results in
@StateT[IdM, Store, Unit]@, and then lift this @StateT@ value to @ReaderT@,
i.e., @AnsM[Unit]@.
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ + αv))
\end{lstlisting}

\paragraph{Primitive Operations}
Other primitive operations can be implemented straightforwardly, for example,
@alloc@ simply takes the size of the current store and produce the successor number.
We elide most of them but describe a little bit on how do we handle functions
and applications as they will be very different when staging and abstract
interpretation is involved. The method @close@ takes a lambda term and an
environment, then create a closure value @CloV@.

@@

To apply a function, the method @ap_clo@ takes a function value and an argument
value, and extracts the lambda term and environment enclosed in the function value,
and evaluates the body expression of that lambda term under the new environment and store.

@@

With other primitive operations implemented, we can implement the top-level
@run@ method, where $\rho_0$ and $\sigma_0$ are the initial empty environment and
store.
